---
title: 'Practical 2: Linear models and ANOVA'
author: "BIO2020"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo = FALSE, message=FALSE,warning= FALSE}
library(dplyr)
library(palmerpenguins)
library(ggformula)

```

For this practical we will be revisitnig the `penguins` data set that we were using in the introductory data visualisation tutorial. 

The `penguins` data are contained in a package called `r emo::ji("package")``palmerpenguins` so rather than importing the data from a `.csv` file we simply need to install and load the package.

**Remember** that the first time we use a package on a computer we need to install it (and we only need to do this once) and that we usually put this code directly in to the console

Once you have installed `r emo::ji("package")``palmerpenguins` you need to load it using the `library()` function

Once the package has been loaded the dataset, which is called `penguins` will be available to use.

Beause the data are associated with the package you will notice that you can interrogate the `penguins object but that it does not appear in your environment pane. 

Using the functions that you learned in the first practical familiarise yourself with the `penguins` data. 

```{r, eval= FALSE}
# see the first 6 rows of the data frame
head(penguins)

# check the dimensiond of the data frame
dim(penguins)

```

For each variable within the dataset note wether it is continous or categoric

 + Species
 + Island
 + Bill length
 + Bill depth
 + Flipper length
 + Body mass
 + Sex
 + Year

## Continous explanatory variables

We will start with a linear model to investigate the relationship between Flipper length and body mass.

Firstly lets plot the data for these two variables, using a scatter plot with our explanatory variable of body mass on the x axis.

```{r echo = FALSE}

gf_point(flipper_length_mm ~ body_mass_g, data = penguins)
```

From this plot can you estimate:

 + Will the intercept of the fitted line be positive or negative?
 + Will the gradient of the fitted line be positive or negative?
 
Let's create a linear model to see if our expectations were correct.

You will remember from the tutorial that the standard syntax for a linear model is 

`lm(response_variable ~ explanatory_variable, data = dataframe_name)`

The function theat we use in R is `lm()` 

**Create a linear model of flipper length and body mass**

**Store the outputs of your linear model as an object called `mass_lm` that you can interrogate**

```{r, echo = FALSE}
mass_lm <- lm(flipper_length_mm~ body_mass_g, data = penguins)
```

Use the `summary()` function to investigate the outputs of your linear model 

```{r}
summary(mass_lm)
```

Were your expectations correct? Using the summary information from your model answer the following questions:

 + What is the intercept of the fitted line?
 + What is the gradient of the fitted line?
 + Is the relationship between the explanatory and response variable significant? 
 + What proportion of the response value can be explained by the explanatory value?  <!--adjusted rsquared--> 
 + Using the `sqrt()` function can you calculate the correlation coefficient between the variables? <!--sqrt(adjusted r2)-->
 + Is the overall model significant?  <!--Fstatistc-->

 
### Plotting the fited line

Add the fitted line from the inear model to your scatterplot using the `gf_lm()` function

Remember that when adding a layer to your plot you will need to use the pipe ` %>% ` which can be created using the keyboard shortcut *Ctrl* + *shift* + *m* (*cmd* + *shift* + *m* on mac)

```{r, echo = TRUE, eval = FALSE}

gf_point(flipper_length_mm ~ body_mass_g, data = penguins) %>% 
        gf_lm()

```

Using the lessons learned in the last practical can you tidy this plot up by: 

+ Removing the grey plot background
+ Altering the axis labels
+ Change the colour of the fitted line

Refer back to your script from the previous practical if you need a reminder of how to do this

```{r echo = FALSE}
gf_point(flipper_length_mm ~ body_mass_g, data = penguins) %>% 
        gf_lm(colour = "#e66101") %>% 
        gf_labs( x = "Body mass (g)", y = "Flipper length (mm)") %>%                
  gf_theme(theme_minimal())
```


### Calculate fitted values for new inputs 

You will remember from the tutorial that the fitted or "predicted" values are those along the straight line that you have plotted using `gf_lm()` and that these values are stored in the model output under the `fitted.values` option which you can access by typing `mass_lm$fitted.values` once you have created your model object.

Using the model that you have creted you need to determine the fitted value for a penguin that weighs **2000g**

Remember that you can use R as a calculator to work out the fitted value using the formula:

`intercept + (gradient * new value)`

or

`intercept - (gradient * new value)`

if the gradient is negative 

Check that your calculation is correct by using the `makeFun()` function  to create a function called `predictor`
The argument to your `makeFun()` function will be the name of your model
Check the predicted value by giving your new `predictor()` function the new input value of 2000g

```{r}
mass_lm <- lm(flipper_length_mm ~ body_mass_g, data = penguins)
predictor <- makeFun(mass_lm)
predictor(2000)
```

Are the values that you obtained the same?
If the values are not exactly the same can you think why that might be? <!--rouding outputs from summary of model-->

### Checking the assumptions of the model 


- check assumptions 
- qq plot - error residuals 


## Categorical explanatory variable

run through same stuff
