---
title: "Summary statistics and Data visualisation"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(mosaic)
library(ggformula)
library(palmerpenguins) #Add install to bio2020 package
library(datasauRus)     #Add install to bio2020 package
#caterpillers <- read.csv("www/regression.csv")
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction

Now that we have gained a throuogh understanding of some of the measures of centrality and variance that we might want to calculate we need to understand how to apply this for our data. 

Here we will run through:

- Generating summary statistics for our dataset (as a whole and for individual variables)
- Visualising and plotting our data

## Checking your data 

Once you have collected your data and read it in to R the first thing to do is to check that you have thr right data!

There are a couple of R functions that offer a useful jumping off point. 
Run the below code to see what output it generates

```{r data-check, exercise=TRUE, exercise.completion=FALSE}
# The head function shows you the first 6 rows of your data set
  head(penguins)
# The names function tells you the names of each of your columns
  names(penguins)
 
```

Now that we have checked our data and we are happy that we are looking at the correct data file we can start to interrogate it a bit further.

We can check the number of rows and columns using the `dim()` function or using `nrow()` or `ncol()` and we can investigate the structure of our data using `str()`

We can use the `summary()` function to give us a summary of the entire dataset which can be a useful way to identify outliers.

Using the summary function can you find out how much the heaviest penguin weighs?

```{r heavy-penguin, exercise=TRUE, exercise.completion=FALSE}

 
```

<div id="heavy-penguin-hint">
**Hint** Use  `args(summary)` to check what inputs the summary function needs 
</div>




